\section{The RationalGRL Tool}
\label{sect:tool}

An important requirement of our framework is that it has tool support. This is for various reasons: i) although there are various approaches attempting to combine goal modeling with argumentation, there does not yet exist any tool support (see Section~\ref{sect:discussion}), ii) it allows us to do user tests in future research, exploring the difference between GRL and RationalGRL, iii) tool support is an excellent verification to ensure our formal framework and algorithms actually work, iv) having a light-weight web-based version of GRL is useful for the community in general.

GRL has a well-documented and well-maintained tool called jUCMNav~\cite{jUCMNav}. This tool is an extension to Eclipse. Although it is a rich tool with many features, it is implemented as an Eclipse plugin. We have experienced on various occasions that setting this up properly can take up a significant time of experiments, leaving less time for actual modeling.

In this section we present our web-based prototype of RationalGRL. It can be accessed from:
\begin{quote}
\url{http://www.rationalgrl.com}
\end{quote}

The tool is written in Javascript, and runs on all modern browsers. The tool is open-source, and the source can be accessed from:
\begin{quote}
\url{http://www.github.com/marcvanzee/RationalGRL}
\end{quote}

In this section we briefly highlight some features of the language, and we explain some of its current limitations.

\todo{all}{marc}{Explain figures in Appendix~\ref{sect:tool-screenshots}}